---
title: "Deriving covariates from gridded ocean data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Deriving covariates from gridded ocean data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4)
library(derivoce)
```

Copernicus gives you temperature, salinity, and currents. What concentrates
plankton is usually not those values but how they are *arranged* and how they are
*changing*: a front rather than a temperature, food accumulated since spring
rather than food present today. This package computes that second kind of
quantity.

Everything here is real output. The vignette builds a small synthetic field so
the examples run without a Copernicus account, and every number and figure below
was produced by the code shown.

## The shape

Every function takes and returns the same thing: an `sf` POINT object with one
row per grid point and time step, a column per covariate, and `YEAR`, `MONTH`,
`DAY` columns. That is what `datamatch::accessEnvDat()` returns.

```{r fetch, eval = FALSE}
env <- datamatch::accessEnvDat(
  vars = c("SST", "SSS", "CHL"), years = 2010, months = 1:6,
  bounding_box = list(xmin = -70, xmax = -66, ymin = 41, ymax = 44)
)
```

Because input and output shapes are identical, the functions compose in a pipe
and nothing has to be reassembled.

For this vignette we build the same shape by hand: a temperature field with a
warm tongue that drifts west through the spring, on a small grid.

```{r field}
lon <- seq(-70, -66, by = 0.2)
lat <- seq(41, 44, by = 0.2)

frames <- lapply(1:6, function(month) {
  grid <- expand.grid(x = lon, y = lat)
  # A warm tongue centred at a longitude that moves west as the season goes on
  centre <- -67.5 - 0.25 * month
  grid$SST <- 8 + 6 * exp(-((grid$x - centre)^2) / 0.6) + 0.35 * (grid$y - 41)
  grid$CHL <- 0.4 + 0.25 * month
  # Fresher inshore, as a coastal current would leave it.
  grid$SSS <- 31 + 0.9 * (grid$x + 70) / 4
  grid$YEAR <- 2010
  grid$MONTH <- month
  grid$DAY <- 1L
  grid
})

env <- sf::st_as_sf(do.call(rbind, frames), coords = c("x", "y"), crs = 4326)
dim(env)
```

## Gradients

`horizontal_gradient()` gives the rate of change with distance, which is what
picks out fronts.

```{r gradient}
env <- horizontal_gradient(env, "SST")
round(range(env$SST_grad, na.rm = TRUE), 3)
```

The units are **covariate units per kilometre**, not per degree. That is the
whole design. A degree of longitude is about 83 km at 42°N and 111 km at the
equator, so a per-degree gradient is stretched by latitude and not comparable
across a study area. Boundary cells are `NA`, because a central difference needs
a neighbour on both sides.

```{r gradient-plot, echo = FALSE}
january <- env[env$MONTH == 1, ]
xy <- sf::st_coordinates(january)
op <- par(mar = c(4, 4, 2, 1))
plot(xy[, 1], january$SST_grad, pch = 16, cex = 0.4, col = "#1f78b4",
     xlab = "longitude", ylab = "SST gradient (°C/km)",
     main = "The front shows up as two peaks")
par(op)
```

Two peaks, because the tongue is warm in the middle: the gradient is steep on
both flanks and flat at the crest. A gradient finds edges, not values.

## Lags and accumulation

Populations respond with a delay, and a survey samples the food built up since
the season began rather than the food present that instant.

```{r temporal}
env <- lag_covariate(env, "SST", n = 1, by = "month")
env <- integrate_covariate(env, "CHL", window = "year")

first_cell <- env[!duplicated(paste(env$MONTH)), ]
data.frame(
  month = first_cell$MONTH,
  CHL = round(first_cell$CHL, 2),
  CHL_int = round(first_cell$CHL_int, 2)
)
```

`CHL_int` is the running sum from January, which is what a copepod integrates.

Note `by = "month"`. Counting *steps* is only unambiguous when the series is
complete: in a monthly record missing April, a one-step lag makes March the
predecessor of May and says nothing about it. Counting calendar months returns
`NA` there instead.

```{r lag-gap}
gappy <- env[env$MONTH != 4, ]

by_step <- lag_covariate(gappy, "SST", n = 1, by = "step")
by_month <- lag_covariate(gappy, "SST", n = 1, by = "month")

c(step = mean(is.na(by_step$SST_lag1)),
  month = mean(is.na(by_month$SST_lag1month)))
```

The calendar version has more `NA` because May genuinely has no April to look
back to. That is the honest answer.

A vector of lags builds an autoregressive set in one call:

```{r ar, eval = FALSE}
env <- lag_covariate(env, "SST", n = 1:3, by = "year")
# adds SST_lag1year, SST_lag2year, SST_lag3year
```

`rolling_covariate()` describes the window instead of accumulating it. The mean
of the last three months and the variability of the last three months are
different covariates, and neither is their sum.

```{r rolling}
env <- rolling_covariate(env, "SST", n = 3, by = "month",
                         stat = c("mean", "sd"))
round(c(mean = mean(env$SST_mean3month, na.rm = TRUE),
        sd = mean(env$SST_sd3month, na.rm = TRUE)), 3)
```

The window is trailing and inclusive, so March covers January to March. Early
steps have less history than the window asks for; `min_obs` decides whether they
get a summary of what is there or an `NA`.

`by = "month"` counts calendar time and `by = "step"` counts positions in the
record. They agree exactly until the series has a gap, and then diverge
silently — which is why the distinction is an argument rather than an
assumption.

## Distance to a front

The local gradient is a blunt predictor of frontal influence. A station in a
smooth patch has zero gradient whether the nearest front is 2 km away or 200 km
away, and those are very different places to be.

```{r fronts}
env <- distance_to_front(env, "SST", quantile = 0.9)
round(summary(env$SST_front_dist), 1)
```

`scope` decides what "a front" means. The default takes one threshold across the
whole record, so a front means the same physical sharpness every month and a
quiet month may contain none. That is a real result rather than something to
suppress: a step with no front returns `NA`, not `0`, because reporting zero
would assert "you are standing on a front".

Fronts also move, and a distance measured at one moment cannot tell a permanent
feature from a passing one. `front_frequency()` asks how *often* a place is
frontal.

```{r front-freq}
env <- front_frequency(env, "SST", scope = "step")
round(summary(env$SST_front_freq), 2)
```

Here the warm tongue drifts west all season, so no cell is frontal for long and
the frequencies stay low. On a real domain the shelf break stands out precisely
because it does not move: a cell frontal in fifteen steps of twenty sits on
something a predator can learn, and one frontal in one caught a filament going
past.

## Anomalies and extremes

Eight degrees is cold for the southern Gulf of Maine and warm for the Scotian
Shelf. A model given raw temperature has to learn that geography before it can
use the departure from it, and `cell_anomaly()` hands the departure over
directly.

```{r anomaly}
env <- cell_anomaly(env, "SST", reference = "record")
round(range(env$SST_anom), 2)
```

`reference = "record"` removes each cell's whole-series mean, which leaves the
seasonal cycle in. `"climatology"` removes a mean per calendar month instead,
which is usually what you want — but it needs several years, because with one
year there is a single January per cell and **every anomaly comes out exactly
zero**. That is a silent failure, so the function warns rather than handing back
a plausible-looking column of zeroes.

`marine_heatwave()` needs the same history, for the same reason: a threshold
taken over three values is the largest of the three. Twelve years of one month
is enough to show it.

```{r heatwave}
years <- 2000:2011
hw <- do.call(rbind, lapply(years, function(y) {
  data.frame(x = -68, y = 43, YEAR = y, MONTH = 1L, DAY = 1L,
             # An ordinary decade, then one very warm year.
             SST = 9 + 0.4 * ((y * 7) %% 5) + if (y == 2011) 6 else 0)
}))
hw <- sf::st_as_sf(hw, coords = c("x", "y"), crs = 4326)

hw <- marine_heatwave(hw, "SST", min_steps = 1)
sf::st_drop_geometry(hw)[hw$mhw_event,
                         c("YEAR", "mhw_intensity", "mhw_category")]
```

One event, in the year it was built into. Intensity is measured from the
climatology rather than from the threshold, so it is comparable between cells
whose thresholds differ, and the category counts how many multiples of the
threshold's own excess the value reached.

## Density

Temperature is a poor stand-in for density wherever salinity varies, and the
Gulf of Maine is such a place: Scotian Shelf inflow arrives cold *and* fresh,
and the two pull density in opposite directions.

```{r density}
env <- potential_density(env)
round(range(env$sigma_theta), 2)
```

That is sigma-theta, density minus 1000, from the UNESCO (1983) equation of
state. Salinity given as a mass fraction rather than PSU sits inside the fitted
range and would return a plausible freshwater density, so a column entirely
below 1 PSU is called out separately.

## Flow structure

With velocities you can compute eddy kinetic energy and Lyapunov exponents. A
small steady field is enough to show the machinery.

```{r flow}
flow <- lapply(1:2, function(month) {
  grid <- expand.grid(x = seq(-69, -68, by = 0.05),
                      y = seq(42, 43, by = 0.05))
  # A westward current with a saddle on top: parcels are carried along, and
  # also separate along one axis while converging on the other.
  grid$UO <- -0.06 + 0.04 * (grid$x + 68.5)
  grid$VO <- -0.04 * (grid$y - 42.5)
  grid$YEAR <- 2010; grid$MONTH <- month; grid$DAY <- 1L
  grid
})
flow <- sf::st_as_sf(do.call(rbind, flow), coords = c("x", "y"), crs = 4326)

flow <- current_speed(flow)
flow <- ftle(flow, integration_days = 5, step_hours = 12)

round(c(speed = median(flow$speed),
        ftle = median(flow$backward_ftle, na.rm = TRUE)), 4)
```

`flow_deformation()` describes the same field without following anything. It
takes the four derivatives of the velocity components and reports what the flow
is doing locally, which costs one pass rather than an integration.

```{r deformation}
flow <- flow_deformation(flow)
measures <- sf::st_drop_geometry(flow)[c("vorticity", "divergence",
                                         "strain_rate", "okubo_weiss")]
signif(sapply(measures, median, na.rm = TRUE), 3)
```

The Okubo–Weiss parameter compares rotation against strain: negative is the
interior of a coherent eddy, positive the filaments between eddies where water
is drawn into long thin structures. This synthetic field is a pure saddle — it
stretches along one axis and squeezes along the other without turning — so
vorticity is zero and Okubo–Weiss is positive throughout.

```{r deformation-sign}
c(rotating = mean(flow$okubo_weiss < 0, na.rm = TRUE),
  straining = mean(flow$okubo_weiss > 0, na.rm = TRUE))
```

The three tools answer different questions about the same velocities.
`eke()` needs a time series and asks how variable a place is.
`flow_deformation()` needs one step and asks what the flow is doing there now.
`ftle()` and `fsle()` need trajectories and ask where water that started apart
ends up together. Cost rises in that order, and so does what they can tell you
about persistence.

`residence_time()` uses the same integrator as `ftle()` to ask how long water
stays in a place.

```{r residence}
box <- list(xmin = -68.8, xmax = -68.2, ymin = 42.2, ymax = 42.8)
flow <- residence_time(flow, box, max_days = 20, step_hours = 12)
round(summary(flow$forward_residence), 1)
```

Points outside the box are `NA`. Anything still inside after `max_days` is
recorded as `max_days`, which is **right-censored**: its real residence is at
least that, and averaging the column biases the answer downwards, most severely
at the most retentive sites. The function warns with the censored fraction for
exactly that reason.

**FTLE is often mostly `NA`, and that is expected.** Parcels are followed through
the velocity field, and one that reaches the edge of the data has nothing left to
follow, so its cell returns `NA`. The cost is a margin of roughly
speed × integration time around the domain. At a shelf speed of 0.15 m/s the
default 14 days is about 180 km, which removes a third of a 500 km box and all of
a small one. Fetch a bounding box larger than your study area by about that
margin. If almost everything comes back `NA`, the function says so and shows the
arithmetic.

## Regional indices

Some quantities describe an area rather than a cell. These return one value per
time step, broadcast to every row, so they behave like a climate index.

```{r transport}
flow <- section_transport(flow, from = c(-68.8, 42.2), to = c(-68.2, 42.8))
tapply(flow$transport, flow$MONTH, function(x) round(unique(x)))
```

One number per month, as intended: the index describes the section, not the cell,
so a horizontal gradient of it would be identically zero and `horizontal_gradient()`
would say so.

Sign follows the direction of travel. Walking the section from `from` to `to`,
flow crossing left to right counts positive, so swapping the endpoints flips it.
Worth checking once against a field whose direction you know, because a sign
error here is invisible: the magnitudes stay plausible and only the
interpretation inverts.

The named cases `scotian_shelf_inflow()` and `northeast_channel_inflow()` are the
same calculation on fixed sections, placed by testing candidate lines against
real GLORYS currents rather than by eye. `docs/methods.md` records how.

`derived_indices()` lists them with the work each follows:

```{r indices}
as.data.frame(derived_indices())[, c("name", "method", "units")]
```

## When the package argues back

Two checks exist because the failure they catch is invisible in the output.

A derivative that cannot carry information gets a warning. Depth does not change
between months, so lagging it returns the column unchanged:

```{r degenerate}
env$DEPTH <- 100 + 20 * (sf::st_coordinates(env)[, 1] + 70)
result <- tryCatch(lag_covariate(env, "DEPTH"),
                   warning = function(w) conditionMessage(w))
cat(substr(result, 1, 180))
```

The check is on the data, not on a list of known column names, so a variable that
happens to be constant in your particular extract is caught too. A horizontal
gradient of `DEPTH` is *not* warned about, because that is slope and a perfectly
sensible covariate. Whether a column belongs depends on the operation, not the
column.

Irregular input is refused outright:

```{r irregular, error = TRUE}
set.seed(1)
scattered <- sf::st_as_sf(
  data.frame(x = runif(50, -70, -66), y = runif(50, 41, 44),
             SST = runif(50), YEAR = 2010, MONTH = 1, DAY = 1L),
  coords = c("x", "y"), crs = 4326
)
horizontal_gradient(scattered, "SST")
```

Gridding scattered points first would produce a gradient field that looks
completely plausible and mostly measures the interpolation.

## Where to go next

`docs/methods.md` in the repository covers how each quantity is computed and why
each choice was made, including the units question behind gradients, the
direction convention for Lyapunov exponents, and how the named sections were
placed and tested. The README summarises the same ground more briefly, with the
reference list.
