---
title: "Deriving covariates from gridded ocean data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Deriving covariates from gridded ocean data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4)
library(derivoce)
```

Copernicus gives you temperature, salinity, and currents. What concentrates
plankton is usually not those values but how they are *arranged* and how they are
*changing*: a front rather than a temperature, food accumulated since spring
rather than food present today. This package computes that second kind of
quantity.

Everything here is real output. The vignette builds a small synthetic field so
the examples run without a Copernicus account, and every number and figure below
was produced by the code shown.

## The shape

Every function takes and returns the same thing: an `sf` POINT object with one
row per grid point and time step, a column per covariate, and `YEAR`, `MONTH`,
`DAY` columns. That is what `datamatch::accessEnvDat()` returns.

```{r fetch, eval = FALSE}
env <- datamatch::accessEnvDat(
  vars = c("SST", "SSS"), years = 2010, months = 1:6,
  bounding_box = list(xmin = -70, xmax = -66, ymin = 41, ymax = 44)
)
```

Because input and output shapes are identical, the functions compose in a pipe
and nothing has to be reassembled.

For this vignette we build the same shape by hand: a temperature field with a
warm tongue that drifts west through the spring, on a small grid.

```{r field}
lon <- seq(-70, -66, by = 0.2)
lat <- seq(41, 44, by = 0.2)

frames <- lapply(1:6, function(month) {
  grid <- expand.grid(x = lon, y = lat)
  # A warm tongue centred at a longitude that moves west as the season goes on
  centre <- -67.5 - 0.25 * month
  grid$SST <- 8 + 6 * exp(-((grid$x - centre)^2) / 0.6) + 0.35 * (grid$y - 41)
  grid$CHL <- 0.4 + 0.25 * month
  grid$YEAR <- 2010
  grid$MONTH <- month
  grid$DAY <- 1L
  grid
})

env <- sf::st_as_sf(do.call(rbind, frames), coords = c("x", "y"), crs = 4326)
dim(env)
```

## Gradients

`horizontal_gradient()` gives the rate of change with distance, which is what
picks out fronts.

```{r gradient}
env <- horizontal_gradient(env, "SST")
round(range(env$SST_grad, na.rm = TRUE), 3)
```

The units are **covariate units per kilometre**, not per degree. That is the
whole design. A degree of longitude is about 83 km at 42°N and 111 km at the
equator, so a per-degree gradient is stretched by latitude and not comparable
across a study area. Boundary cells are `NA`, because a central difference needs
a neighbour on both sides.

```{r gradient-plot, echo = FALSE}
january <- env[env$MONTH == 1, ]
xy <- sf::st_coordinates(january)
op <- par(mar = c(4, 4, 2, 1))
plot(xy[, 1], january$SST_grad, pch = 16, cex = 0.4, col = "#1f78b4",
     xlab = "longitude", ylab = "SST gradient (°C/km)",
     main = "The front shows up as two peaks")
par(op)
```

Two peaks, because the tongue is warm in the middle: the gradient is steep on
both flanks and flat at the crest. A gradient finds edges, not values.

## Lags and accumulation

Populations respond with a delay, and a survey samples the food built up since
the season began rather than the food present that instant.

```{r temporal}
env <- lag_covariate(env, "SST", n = 1, by = "month")
env <- integrate_covariate(env, "CHL", window = "year")

first_cell <- env[!duplicated(paste(env$MONTH)), ]
data.frame(
  month = first_cell$MONTH,
  CHL = round(first_cell$CHL, 2),
  CHL_int = round(first_cell$CHL_int, 2)
)
```

`CHL_int` is the running sum from January, which is what a copepod integrates.

Note `by = "month"`. Counting *steps* is only unambiguous when the series is
complete: in a monthly record missing April, a one-step lag makes March the
predecessor of May and says nothing about it. Counting calendar months returns
`NA` there instead.

```{r lag-gap}
gappy <- env[env$MONTH != 4, ]

by_step <- lag_covariate(gappy, "SST", n = 1, by = "step")
by_month <- lag_covariate(gappy, "SST", n = 1, by = "month")

c(step = mean(is.na(by_step$SST_lag1)),
  month = mean(is.na(by_month$SST_lag1month)))
```

The calendar version has more `NA` because May genuinely has no April to look
back to. That is the honest answer.

A vector of lags builds an autoregressive set in one call:

```{r ar, eval = FALSE}
env <- lag_covariate(env, "SST", n = 1:3, by = "year")
# adds SST_lag1year, SST_lag2year, SST_lag3year
```

## Distance to a front

The local gradient is a blunt predictor of frontal influence. A station in a
smooth patch has zero gradient whether the nearest front is 2 km away or 200 km
away, and those are very different places to be.

```{r fronts}
env <- distance_to_front(env, "SST", quantile = 0.9)
round(summary(env$SST_front_dist), 1)
```

`scope` decides what "a front" means. The default takes one threshold across the
whole record, so a front means the same physical sharpness every month and a
quiet month may contain none. That is a real result rather than something to
suppress: a step with no front returns `NA`, not `0`, because reporting zero
would assert "you are standing on a front".

## Flow structure

With velocities you can compute eddy kinetic energy and Lyapunov exponents. A
small steady field is enough to show the machinery.

```{r flow}
flow <- lapply(1:2, function(month) {
  grid <- expand.grid(x = seq(-69, -68, by = 0.05),
                      y = seq(42, 43, by = 0.05))
  # A westward current with a saddle on top: parcels are carried along, and
  # also separate along one axis while converging on the other.
  grid$UO <- -0.06 + 0.04 * (grid$x + 68.5)
  grid$VO <- -0.04 * (grid$y - 42.5)
  grid$YEAR <- 2010; grid$MONTH <- month; grid$DAY <- 1L
  grid
})
flow <- sf::st_as_sf(do.call(rbind, flow), coords = c("x", "y"), crs = 4326)

flow <- current_speed(flow)
flow <- ftle(flow, integration_days = 5, step_hours = 12)

round(c(speed = median(flow$speed),
        ftle = median(flow$backward_ftle, na.rm = TRUE)), 4)
```

**FTLE is often mostly `NA`, and that is expected.** Parcels are followed through
the velocity field, and one that reaches the edge of the data has nothing left to
follow, so its cell returns `NA`. The cost is a margin of roughly
speed × integration time around the domain. At a shelf speed of 0.15 m/s the
default 14 days is about 180 km, which removes a third of a 500 km box and all of
a small one. Fetch a bounding box larger than your study area by about that
margin. If almost everything comes back `NA`, the function says so and shows the
arithmetic.

## Regional indices

Some quantities describe an area rather than a cell. These return one value per
time step, broadcast to every row, so they behave like a climate index.

```{r transport}
flow <- section_transport(flow, from = c(-68.8, 42.2), to = c(-68.2, 42.8))
tapply(flow$transport, flow$MONTH, function(x) round(unique(x)))
```

One number per month, as intended: the index describes the section, not the cell,
so a horizontal gradient of it would be identically zero and `horizontal_gradient()`
would say so.

Sign follows the direction of travel. Walking the section from `from` to `to`,
flow crossing left to right counts positive, so swapping the endpoints flips it.
Worth checking once against a field whose direction you know, because a sign
error here is invisible: the magnitudes stay plausible and only the
interpretation inverts.

The named cases `scotian_shelf_inflow()` and `northeast_channel_inflow()` are the
same calculation on fixed sections, placed by testing candidate lines against
real GLORYS currents rather than by eye. `docs/methods.md` records how.

`derived_indices()` lists them with the work each follows:

```{r indices}
as.data.frame(derived_indices())[, c("name", "method", "units")]
```

## When the package argues back

Two checks exist because the failure they catch is invisible in the output.

A derivative that cannot carry information gets a warning. Depth does not change
between months, so lagging it returns the column unchanged:

```{r degenerate}
env$DEPTH <- 100 + 20 * (sf::st_coordinates(env)[, 1] + 70)
result <- tryCatch(lag_covariate(env, "DEPTH"),
                   warning = function(w) conditionMessage(w))
cat(substr(result, 1, 180))
```

The check is on the data, not on a list of known column names, so a variable that
happens to be constant in your particular extract is caught too. A horizontal
gradient of `DEPTH` is *not* warned about, because that is slope and a perfectly
sensible covariate. Whether a column belongs depends on the operation, not the
column.

Irregular input is refused outright:

```{r irregular, error = TRUE}
set.seed(1)
scattered <- sf::st_as_sf(
  data.frame(x = runif(50, -70, -66), y = runif(50, 41, 44),
             SST = runif(50), YEAR = 2010, MONTH = 1, DAY = 1L),
  coords = c("x", "y"), crs = 4326
)
horizontal_gradient(scattered, "SST")
```

Gridding scattered points first would produce a gradient field that looks
completely plausible and mostly measures the interpolation.

## Where to go next

`docs/methods.md` in the repository covers how each quantity is computed and why
each choice was made, including the units question behind gradients, the
direction convention for Lyapunov exponents, and how the named sections were
placed and tested. The README summarises the same ground more briefly, with the
reference list.
